Beyond Capacity: Analyzing Shelter Use and Government Responses to Homelessness in Toronto

### 2024-03-16 paper/paper.qmd
new updatessss

During the COVID-19 pandemic, urban shelters faced unprecedented challenges. The pandemic's ripple effects have reached Toronto's shelters, which provide a safety net for the city's homeless population. As public health directives evolved and governmental policies changed, the shelter system's response became a critical area of research to ensure the homeless's safety and well-being during and after the crisis.
This paper looks at the complexities of shelter use, service demand, and capacity in Toronto's diverse urban landscape in 2022. Despite previous research on homelessness, there are few comprehensive analyses that combine a Bayesian approach with a visual examination of service patterns across shelter types and demographics. This study seeks to close that gap by providing a nuanced analysis of the current state of the shelter system using sophisticated statistical models and extensive visual data. On the other hand, we discovered some secondary resources about the city of Toronto, as well as any other pertinent resources. According to a City of Toronto article, the city plans to build over 400 affordable homes for homeless people, with 1,100 supportive homes completed in 2020. By 2022, the City of Toronto expects nearly 3,000 new housing opportunities.However, we also cite the article "Prevalence of SARS-CoV-2 infection among people experiencing homelessness in Toronto during the first wave of the COVID-19 pandemic" by the author Linh Luong and others that The COVID-19 pandemic has had a significant impact on homelessness, especially among Indigenous and racialized communities. Over 35,000 people in Canada are homeless on a daily basis, with Toronto's homeless population being especially diverse.

@@

These discoveries are undeniably significant. The high proportion of men in shelter occupancy data necessitates a thorough reassessment of existing support structures designed specifically for men. It requires that resource allocation and policymaking be gender-sensitive. The high turnover rates and seasonal peaks in occupancy among men highlight the need for targeted interventions that address not only immediate shelter needs but also more general systemic issues such as employment, mental health, and substance abuse—all of which are frequently associated with homelessness. By highlighting these issues, our research contributes to a more informed conversation, which is critical for developing strong, long-term solutions for Toronto's homeless men.

Our method, which we use We created and analyzed the statistic with the R. R Core Team (2023) and Wickham et al. (2019). This study employs Bayesian statistical models to examine shelter service data from across the city, including occupancy rates and service user demographics. Multiple bar charts and other graphical representations demonstrate the temporal and spatial dynamics of shelter use, revealing underlying trends and informing model parameters.

# Data {#sec-data}

From @tbl-torontoo,

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in.

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

prior_summary(toronto_shelters_data_first_model_rstanarm)
```

@@

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

we implement a posterior predictive check. This shows...
we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"

pp_check(toronto_shelters_data_first_model_rstanarm) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Diagnostics

is a trace plot. It shows... This suggests...

is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(toronto_shelters_data_first_model_rstanarm, "trace")

plot(toronto_shelters_data_first_model_rstanarm, "rhat")
```

>>>>

During the COVID-19 pandemic, urban shelters faced unprecedented challenges. The pandemic's ripple effects have reached Toronto's shelters, which provide a safety net for the city's homeless population. As public health directives evolved and governmental policies changed, the shelter system's response became a critical area of research to ensure the homeless's safety and well-being during and after the crisis.
This paper looks at the complexities of shelter use, service demand, and capacity in Toronto's diverse urban landscape in 2022. Despite previous research on homelessness, there are few comprehensive analyses that combine a Bayesian approach with a visual examination of service patterns across shelter types and demographics. This study seeks to close that gap by providing a nuanced analysis of the current state of the shelter system using sophisticated statistical models and extensive visual data. On the other hand, we discovered some secondary resources about the city of Toronto, as well as any other pertinent resources. According to a City of Toronto article, the city plans to build over 400 affordable homes for homeless people, with 1,100 supportive homes completed in 2020. By 2022, the City of Toronto expects nearly 3,000 new housing opportunities.However, we also cite the article "Prevalence of SARS-CoV-2 infection among people experiencing homelessness in Toronto during the first wave of the COVID-19 pandemic" by the author Linh Luong[@luong2022prevalence] and others that The COVID-19 pandemic has had a significant impact on homelessness, especially among Indigenous and racialized communities. Over 35,000 people in Canada are homeless on a daily basis, with Toronto's homeless population being especially diverse.

@@

These discoveries are undeniably significant. The high proportion of men in shelter occupancy data necessitates a thorough reassessment of existing support structures designed specifically for men. It requires that resource allocation and policy making be gender-sensitive. The high turnover rates and seasonal peaks in occupancy among men highlight the need for targeted interventions that address not only immediate shelter needs but also more general systemic issues such as employment, mental health, and substance abuse—all of which are frequently associated with homelessness. By highlighting these issues, our research contributes to a more informed conversation, which is critical for developing strong, long-term solutions for Toronto's homeless men.
# Data {#sec-data}

The data set was collected by the city of Toronto's data @opendatatoronto. We have choose the daily shelter and overnight service occupancy and capacity as the main direction of research. We analysis the data in R(@citeR), with additional tools for support my analysis, including `tidyverse`[@citetidyverse], `here`[@citehere], `dplyr`[@citedplyr], `readr`[@citereadr], `modelsummary`[@citemodelsummary], `janitor`[@citejanitor], `tibble`[@citetibble], `ggplot2`[@citeggplot2], and research method is from @tellingstorieswithdata.

@@

In our study, the study used a Bayesian analysis model to investigate occupancy patterns and service utilization in Toronto's shelter system in 2022. The model was chosen because of its ability to account for prior knowledge and uncertainty, resulting in a probabilistic approach to understanding shelter usage data. The model used a hierarchical structure to account for the different effects of factors like location, demographic groups, and time of year on shelter occupancy rates. The posterior distributions of model parameters revealed information about the relationships between these factors and shelter usage, particularly across different parts of the city and demographic groups.

However, The Bayesian model used in the analysis was effective in drawing significant conclusions about Toronto's shelter service usage trends. The study provided a multidimensional view of the shelter system by combining quantitative analysis and data visualization, as well as identifying significant trends and patterns that can inform service delivery and policy. Instead of being based on an image, the information about the priors being normally distributed with a location of 0 and a scale of 2.5 is based on a standard Bayesian statistics method for defining model priors. A normal distribution is frequently used as a weakly informative prior. This option represents a baseline assumption that, prior to data observation, the coefficients should be centered around 0, indicating no effect with a degree of uncertainty represented by the scale.

The model specified occupancy rates as a function of several covariates, including time (months), city location (e.g., Etobicoke, North York, Scarborough, Toronto, Vaughan), and shelter demographics. Specifically, we used the number of occupied rooms (number_occupied) as a response variable, with (occupancy_month) and (location_city) as predictor variables, as well as sectors to account for different demographic groups.

@@

$y_i | \mu_i \sim \text{Normal}(\mu_i, \sigma)$: Following the Bayesian model, we know that this implies that the outcome variable $y_i$, conditioned on the mean $\mu_i$, has a normal distribution with standard deviation $\sigma$. For the Toronto shelter, $y_i$ can represent the number of people who come to the shelter on day $i$, or other relevant response metric, such as the number of occupied beds.

$\mu_i = \beta_0 + \beta_1 \cdot x_i$ : This is the linear model, where $\mu_i$ represents the expected value of $y_i$. The intercept $\beta_0$ and slope $beta_1$ are multiplied by the predictor $x_i$. In the Toronto shelter, $x_i$ represents a variable of interest, it  could represent the variables such as the capacity of the shelter, the number of services offered, location factors, or the time of year.

$\beta_0, \sim \text{Normal}(0, 2.5)$ : The intercept term $\beta_0$ is assumed to be centred around 0 with a standard deviation of 2.5. This means that you do not expect a strong relationship between the predictor variable $x_i$ and the outcome $y_i$ before analyzing the data, but you are open to the possibility of discovering one.This expresses modest prior knowledge about shelter occupancy rates, allowing the data to significantly influence the posterior distribution $\beta_0$.

$\beta_1 \sim \text{Normal}(0, 2.5)$ : In a comparable way, this indicates the previous consensus regarding the slope β1 was that it has a standard deviation of 2.5 and is centered around 0. This implies that, prior to data analysis, you do not expect a strong relationship between the predictor variable $x_i$ and the outcome of $y_i$.

$\sigma \sim \text{Exponential}(1)$ : This implies that the prior belief for the standard deviation of the outcome variable's distribution is an exponential distribution with a rate parameter of one. This suggests lower σ values, indicating less variability in daily shelter use. However, you are open to learning from the data if there is more variability than expected.This distribution is commonly used for scale parameters like σ in Bayesian models because it is constrained to positive values and can represent a wide range of variability.

Furthermore, the Bayesian model is used in our study to examine occupancy rates in Toronto's shelter system, allowing for a thorough understanding of the uncertainties and variables associated with shelter use. This method generates preliminary estimates, which can be refined as more data becomes available, resulting in a probabilistic forecast. The model's predictive power stems from its ability to differentiate between the effects of various factors on the number of people seeking shelter. By assigning priors to model parameters and tracking their evolution through posterior distributions, the model learns about the relative weight of each factor, allowing it to identify which signals are genuine and worth paying attention to.
The Bayesian framework allows for a dynamic dialogue between data and theory, challenging and refining our understanding of the relationship between socioeconomic factors and shelter demand. This dialogue lays the groundwork for developing targeted policies that are both responsive to immediate needs and proactive in promoting long-term solutions to homelessness. The model's findings serve as an empirical foundation for resource allocation discussions, allowing policymakers to prioritize funding, tailor services to demographic needs, and design interventions that directly address identified causes of shelter overuse.

@@

The Bayesian model was used to analyzing in Toronto's shelter occupancy rates since it is reliable, adaptable, and possesses a lot of information. The model analyzes what it currently knows and adds it to the data it has seen to make an accurate assessment that is based on both past events and present conditions. It finds possible correlations in nested data, which lets you get an accurate understanding of how individual and group-level covariates affect occupancy rates. Thus, the Bayesian approach measures uncertainty by giving a full probabilistic account of all unknowns.

However, the model are also flexible, since it can be updated all the time as new data comes in. This makes it a dynamic tool for ongoing policy change and assessment. The Bayesian model also lets us examine at the complicated connections and interactions between covariates like demographic variables, economic factors, and neighbourhood characteristics. It Figuring out these connections can help you see deeper patterns and trends that other modelling methods might miss. Lastly, the Bayesian model was picked because it is reliable, adaptable, and gives a lot of information. This makes it an important tool for understanding the Toronto shelter system and making it work better and have a bigger effect.

@@

@@
